Component response rate variation drives stability in large complex systems

### 2018-08-15 notebook/ms.Rmd
Updates for PLoS Comp Biol -- new Servan et al reference add

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: yes
linkcolor: blue
csl: plos.csl
biblio-style: apalike
---

@@

**The stability of a complex system generally decreases with increasing system size, as is demonstrated by random matrix theory [@May1972; @Allesina2012]. This counter-intuitive result, first shown by May [@May1972], is broadly relevant for understanding the dynamics and persistence of systems such as ecological [@May1972; @Allesina2012], neurological [@Gray2008; @Gray2009], biochemical [@Rosenfeld2009; @MacArthur2010] and socio-economic [@Haldane2011; @Suweis2014; @Bardoscia2017] networks. Much attention has especially been given to the stability of ecological communities such as food webs or mutualist networks, with recent work investigating how different community structures affect stability [@Allesina2012; @Mougi2012; @Allesina2015a; @Gao2016; @Grilli2017; @Patel2018]. But more broadly, stabilising mechanisms in complex systems remain under-developed, and the effect of variation in the response rate of individual system components remains an open problem [@Allesina2015]. Here I show that when components of a complex system respond to system perturbation at different rates ($\boldsymbol{\gamma}$), the potential for system stability is markedly increased. Stability is caused by the clustering of some eigenvalues toward the centre of eigenvalue distributions despite the destabilising effect of higher interaction strength variation ($\boldsymbol{\sigma^{2}}$). This effect of variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, to the extent that the largest stable complex systems would otherwise be unstable if not for $\boldsymbol{Var(\gamma)}$. My results therefore reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems. Future research in complex systems should therefore account for the varying response rates of individual system components when assessing whole system stability.**

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that the system is stable and values of all components are positive at equilibrium [@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic [@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities [@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

@@

Interestingly, while complex systems were more likely to be stable given variation in component response rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species [@Stone2017; @Grilli2017; @Dougoud2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the stability of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems [@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

@@

<!--- USE J INSTEAD TO CLARIFY M --->

**Component response rate variation ($\mathbf{\gamma}$)**. In a synthesis of eco-evolutionary feedbacks on community stability, Patel et al. model a system that includes a vector of potentially changing species densities ($\mathbf{N}$) and a vector of potentially evolving traits ($\mathbf{x}$)[@Patel2018]. For any species $i$ or trait $j$, change in species density ($N_{i}$) or trait value ($x_{j}$) with time ($t$) is a function of the vectors $\mathbf{N}$ and $\mathbf{x}$,

@@

In matrix notation[@May1973],

@@

**System feasibility**. Dougoud et al. [@Dougoud2018] define the following feasibility criteria for ecological systems characterised by $S$ interacting species with varying densities in a classical Lotka-Volterra model,

@@

In the above, $\mathbf{x^{*}}$ is the vector of species densities at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is the strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion [@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below,

@@

If more discussion is needed, then talk about relation to Jirsa and Ding 2004.

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: yes
linkcolor: blue
csl: plos.csl
biblio-style: apalike
---

@@

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that the system is stable and values of all components are positive at equilibrium [@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic [@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities [@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$ and instead occurred with a fixed probability of $1/2^{S}$, consistent with a recent proof by Serv&aacute;n et al. [@Servan2018] (see Supplementary Information). Hence, for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

@@

Interestingly, while complex systems were more likely to be stable given variation in component response rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species [@Stone2017; @Grilli2017; @Dougoud2018; @Servan2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the feasibility of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems [@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

@@

**Component response rate variation ($\mathbf{\gamma}$)**. In a synthesis of eco-evolutionary feedbacks on community stability, Patel et al. model a system that includes a vector of potentially changing species densities ($\mathbf{N}$) and a vector of potentially evolving traits ($\mathbf{x}$) [@Patel2018]. For any species $i$ or trait $j$, change in species density ($N_{i}$) or trait value ($x_{j}$) with time ($t$) is a function of the vectors $\mathbf{N}$ and $\mathbf{x}$,

@@

In matrix notation [@May1973],

@@

**System feasibility**. Dougoud et al. [@Dougoud2018] identify the following feasibility criteria for ecological systems characterised by $S$ interacting species with varying densities in a classical Lotka-Volterra model,

@@

In the above, $\mathbf{x^{*}}$ is the vector of species densities at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is the strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's [@May1972] stability criterion [@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below (see also [@Servan2018]),

@@

### 2018-08-15 notebook/ms.tex
Updates for PLoS Comp Biol -- new Servan et al reference add

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at equilibrium
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}{]}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically realistic
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]}. While my results are
intended to be general to all complex systems, and not restricted to
species networks, I have also performed a feasibility analysis on all
matrices \(\mathbf{M}\) tested for stability, and additionally for
specific types of ecological communities
{[}\protect\hyperlink{ref-Allesina2012}{2}{]} (e.g., competitive,
mutualist, predator-prey; see Supplementary Information). Feasibility
was unaffected by \(Var(\gamma)\), meaning that for pure interacting
species networks, variation in component response rate (i.e., species
generation time) does not affect stability at biologically realistic
species densities.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of species
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20}{]}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the stability of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical systems {[}\protect\hyperlink{ref-Patel2018}{15}{]}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits (\(\mathbf{x}\)){[}\protect\hyperlink{ref-Patel2018}{15}{]}. For
any species \(i\) or trait \(j\), change in species density (\(N_{i}\))
or trait value (\(x_{j}\)) with time (\(t\)) is a function of the
vectors \(\mathbf{N}\) and \(\mathbf{x}\),

@@

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-May1973}{22}{]},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(a_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation{[}\protect\hyperlink{ref-May1973}{22}{]},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} =\mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{22}{]}.

@@

\textbf{System feasibility}. Dougoud et al.
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]} define the following
feasibility criteria for ecological systems characterised by \(S\)
interacting species with varying densities in a classical Lotka-Volterra
model,

@@

In the above, \(\mathbf{x^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{x^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's{[}\protect\hyperlink{ref-May1972}{1}{]} stability
criterion {[}\protect\hyperlink{ref-Dougoud2018}{20}{]} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below,

@@

\hypertarget{ref-May1973}{}
22. May RM. Qualitative stability in model ecosystems. Ecology. 1973;54:
638--641. doi:\href{https://doi.org/10.2307/1935352}{10.2307/1935352}

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.725 and 0.741 for a and b,
respectively.

>>>>

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at equilibrium
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}{]}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically realistic
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]}. While my results are
intended to be general to all complex systems, and not restricted to
species networks, I have also performed a feasibility analysis on all
matrices \(\mathbf{M}\) tested for stability, and additionally for
specific types of ecological communities
{[}\protect\hyperlink{ref-Allesina2012}{2}{]} (e.g., competitive,
mutualist, predator-prey; see Supplementary Information). Feasibility
was unaffected by \(Var(\gamma)\) and instead occurred with a fixed
probability of \(1/2^{S}\), consistent with a recent proof by Serván et
al. {[}\protect\hyperlink{ref-Servan2018}{22}{]} (see Supplementary
Information). Hence, for pure interacting species networks, variation in
component response rate (i.e., species generation time) does not affect
stability at biologically realistic species densities.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of species
{[}\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Servan2018}{22}{]}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the feasibility of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical systems {[}\protect\hyperlink{ref-Patel2018}{15}{]}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

@@

\textbf{Component response rate variation (\(\mathbf{\gamma}\))}. In a
synthesis of eco-evolutionary feedbacks on community stability, Patel et
al. model a system that includes a vector of potentially changing
species densities (\(\mathbf{N}\)) and a vector of potentially evolving
traits (\(\mathbf{x}\)) {[}\protect\hyperlink{ref-Patel2018}{15}{]}. For
any species \(i\) or trait \(j\), change in species density (\(N_{i}\))
or trait value (\(x_{j}\)) with time (\(t\)) is a function of the
vectors \(\mathbf{N}\) and \(\mathbf{x}\),

@@

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-May1973}{23}{]},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(a_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation {[}\protect\hyperlink{ref-May1973}{23}{]},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} =\mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of May
{[}\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{23}{]}.

@@

\textbf{System feasibility}. Dougoud et al.
{[}\protect\hyperlink{ref-Dougoud2018}{20}{]} identify the following
feasibility criteria for ecological systems characterised by \(S\)
interacting species with varying densities in a classical Lotka-Volterra
model,

@@

In the above, \(\mathbf{x^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{x^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's {[}\protect\hyperlink{ref-May1972}{1}{]} stability
criterion {[}\protect\hyperlink{ref-Dougoud2018}{20}{]} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below (see also
{[}\protect\hyperlink{ref-Servan2018}{22}{]}),

@@

\hypertarget{ref-Servan2018}{}
22. Serván CA, Capitán JA, Grilli J, Morrison KE, Allesina S.
Coexistence of many species in random ecosystems. Nature Ecology and
Evolution. 2018;2: 1237--1242.
doi:\href{https://doi.org/10.1038/s41559-018-0603-6}{10.1038/s41559-018-0603-6}

\hypertarget{ref-May1973}{}
23. May RM. Qualitative stability in model ecosystems. Ecology. 1973;54:
638--641. doi:\href{https://doi.org/10.2307/1935352}{10.2307/1935352}

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.721 and 0.739 for a and b,
respectively.
